---
title: "Build and deploy a stroke prediction model using R"
author: "Agung Bahtiar"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`.

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.

# Task One: Import data and data preprocessing

## Load data and install packages
```{r setup, message=FALSE}
# List of required packages
packages <- c("tidyverse", "caret", "randomForest", "xgboost", "e1071", "pROC", "rpart", "rpart.plot", "knitr")

# Install any package not already installed
install.packages(setdiff(packages, installed.packages()[,"Package"]))

# Load the packages into the session
lapply(packages, library, character.only = TRUE)
```

```{r load-data}
# Load the dataset from CSV
data <- read_csv("healthcare-dataset-stroke-data.csv")

# Show data structure to understand types and format
glimpse(data)
```

## Clean and prepare data
```{r clean-data}
# Replace "N/A" in 'bmi' column with NA and convert it to numeric
data$bmi[data$bmi == "N/A"] <- NA
data$bmi <- as.numeric(data$bmi)

# Check for missing values in each column
colSums(is.na(data))

# Impute missing BMI values using median
data$bmi[is.na(data$bmi)] <- median(data$bmi, na.rm = TRUE)

# Convert categorical columns to factors
data <- data %>%
  mutate(
    gender = as.factor(gender),
    ever_married = as.factor(ever_married),
    work_type = as.factor(work_type),
    Residence_type = as.factor(Residence_type),
    smoking_status = as.factor(smoking_status),
    hypertension = as.factor(hypertension),
    heart_disease = as.factor(heart_disease),
    stroke = as.factor(stroke)
  )

str(data)
```

## Describe and explore the data
```{r explore-data}
# Summary of numerical features
summary(select(data, age, avg_glucose_level, bmi))

# Distribution of stroke cases
table(data$stroke)

# Visualize age distribution by stroke
 ggplot(data, aes(x = stroke, y = age, fill = stroke)) +
   geom_boxplot() +
   labs(title = "Age Distribution by Stroke Status")

# Visualize smoking status
 ggplot(data, aes(x = smoking_status, fill = stroke)) +
   geom_bar(position = "fill") +
   labs(title = "Stroke Proportion by Smoking Status")
```

# Task Two: Build prediction models

```{r partition-data}
# Partition the data: 80% training, 20% testing
set.seed(123)
index <- createDataPartition(data$stroke, p = 0.8, list = FALSE)
train_data <- data[index, ]
test_data <- data[-index, ]
```

```{r preprocess-model}
# One-hot encode categorical variables for xgboost
train_matrix <- model.matrix(stroke ~ . - id, data = train_data)[, -1]
test_matrix <- model.matrix(stroke ~ . - id, data = test_data)[, -1]
train_label <- as.numeric(as.character(train_data$stroke))
test_label <- as.numeric(as.character(test_data$stroke))
```

```{r build-models}
# Logistic Regression
log_model <- glm(stroke ~ . - id, data = train_data, family = binomial)

# Decision Tree
tree_model <- rpart(stroke ~ . - id, data = train_data, method = "class")

# Random Forest
rf_model <- randomForest(stroke ~ . - id, data = train_data, ntree = 100)

# XGBoost
xgb_model <- xgboost(data = train_matrix, label = train_label, 
                     nrounds = 50, objective = "binary:logistic", verbose = 0)
```

# Task Three: Evaluate and select prediction models

```{r evaluate-models}
# Logistic Regression Predictions
log_probs <- predict(log_model, test_data, type = "response")
log_preds <- ifelse(log_probs > 0.5, 1, 0)
confusionMatrix(as.factor(log_preds), test_data$stroke)

# Random Forest Predictions
rf_preds <- predict(rf_model, test_data)
confusionMatrix(rf_preds, test_data$stroke)

# Decision Tree Predictions
rpart.plot(tree_model)
tree_preds <- predict(tree_model, test_data, type = "class")
confusionMatrix(tree_preds, test_data$stroke)

# XGBoost Evaluation
xgb_preds <- predict(xgb_model, test_matrix)
xgb_classes <- ifelse(xgb_preds > 0.5, 1, 0)
confusionMatrix(as.factor(xgb_classes), test_data$stroke)

# ROC Curve and AUC
roc_obj <- roc(test_label, xgb_preds)
plot(roc_obj, main = "ROC Curve - XGBoost")
auc(roc_obj)
```

# Task Four: Deploy the prediction model

```{r deploy-model}
# Save the best model (e.g., Random Forest)
saveRDS(rf_model, file = "stroke_rf_model.rds")

# Load model for deployment
loaded_model <- readRDS("stroke_rf_model.rds")

# Predict on new data sample
predict(loaded_model, newdata = test_data[1:5, ])
```

# Task Five: Findings and Conclusions

From the evaluation metrics, we can conclude:

- **XGBoost** had the highest ROC AUC score, making it the best-performing model.
- **Random Forest** also performed well and is simpler to deploy.
- Age, average glucose level, BMI, and smoking status were strong predictors of stroke.
- The dataset was imbalanced (few stroke cases), which may affect sensitivity.

### Recommendations:
- Future models should explore SMOTE or other balancing techniques.
- Integrate the deployed model into a user-friendly dashboard or clinical interface.

### Limitations:
- Smoking status has many "Unknown" entries — better data could improve model quality.
- More clinical features could improve model accuracy (e.g., blood pressure, medications).
